String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-10-19 2023-1-19/sections/algorithms.tex
remove some contents
section 6

This section aims to show how to solve RECL constraints by utilizing CEFA. 
At first, we reduce the satisfiability of RECL constraints to a decision problem defined in the following section. Then we propose an decision procedure to solve the problem.

\begin{definition}[Nonemptiness of CEFA w.r.t. an LIA formula
, abbreviated as $\cefadec$]
Suppose that $x_1, \cdots, x_n$ are string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, and the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, moreover, $\varphi$ is an LIA formula whose free variables are from $ \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are assignments $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{i,j}}$
such that $(\theta(x_i); \myvec{v_{i,j}}) \in \Lang(\aut_{i, j})$ and  $\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})]$ is true, for every $i \in [n], j \in [l_i]$.
\end{definition}
\begin{proposition}\label{prop-cefadec}
$\cefadec$ is PSPACE-complete. 
\end{proposition}
\begin{proof}
  The lower bound follows from the fact that the nonemptiness problem of the intersection of NFAs is PSPACE-complete \cite{pspace_int}. For the upper bound, \cite{atva2020} showed that $\cefadec$ is in PSPACE by reducing it to monotonic counter machines.
\end{proof}
Note that the decision procedure in \cite{atva2020} was only used to prove the upper bound in Proposition~\ref{prop-cefadec} and not implemented as a matter of fact. Instead, the symbolic model checker nuXmv was used to solve $\cefadec$. We do not rely on nuXmv in this work and shall propose a new algorithm for solving $\cefadec$ in Section~\ref{subsec:cefadec}.

@@

\subsection{Solving $\cefadec$} \label{subsec:cefadec}

@@

\medskip
\noindent
\emph{Step 1.(Intersection)} For each $i \in [n]$, compute a CEFA $\cB_{i} = \aut_{i,1} \cap \cdots \cap \aut_{i, l_i}$, and let $\Lambda'_{x_i} := \{\cB_i\}$. \qed

After Step 1, the nonemptiness of CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi$ is reduced to the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$.

In the following steps, we reduce the non-emptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$ to the satisfiability of an LIA formula. The reduction relies on two observations: The first observation follows that $\Lambda'_{x_i}$ for each $i \in [n]$ contains exactly one CEFA, and the second observation follows that the non-emptiness of exactly one CEFA w.r.t LIA formulae is entirely unrelated to the alphabet.

@@

\begin{observation}\label{obs-cefa-nfa}
For each CEFA $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$, 
%
$$\cefaout(\aut) = \left\{\sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \  \big{|}\ \eta \in \parikh(\cU_\aut) \mbox{ and }  \alpha\left[ \sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \big{/} R \right] \mbox{ is true}\right\}.$$
%
\end{observation}

@@

Because the sizes of the LIA formulas $\psi_{\cU_{\cB_1}}(\fZ_{\costset_{\cB_1}}), \cdots, \psi_{\cU_{\cB_n}}(\fZ_{\costset_{\cB_n}})$ are proportional to the sizes (more precisely, the alphabet size, the number of states and transitions) of NFA $\cU_{\cB_1}, \cdots, \cU_{\cB_n}$, and the satisfiability of existential LIA formulas is NP-complete, it is vital to reduce the sizes of these NFA to improve the performance.

Since
$\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v} = \sum \limits_{\myvec{v} \in \costset(\cB_i) \setminus \{\myvec{0}\}} \eta(\myvec{v}) \myvec{v}$ for each $i \in [n]$ and $\eta \in \parikh(\cU_{\cB_i})$, it turns out that 
the $\myvec{0}$-labeled transitions in $\cU_{\cB_i}$ do not contribute to the final output $\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v}$. Therefore, we propose the following size-reduction technique for the NFA $\cU_{\cB_i}$'s. Regardless of alphabet, we take register updates as new alphabet symbols and $\myvec{0}$ as $\epsilon$ symbol. Then we are capable of using standard minimization to reduce the automata size.

\medskip
\noindent
\emph{Step 2.(Automata size reduction)} For each $i \in [n]$, we view the transitions $(q, \myvec{0}, q')$ in $\cU_{\cB_i}$ as $\varepsilon$-transitions $(q, \varepsilon, q')$, and remove the $\varepsilon$-transitions from $\cU_{\cB_i}$. Then we determinize and minimize the resulting NFA.   \qed

@@

\medskip
\noindent
\emph{Step 3.(Parikh image)} For each $i \in [n]$, we compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cC_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then we solve the satisfiability of the following formula,
%
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
\qed

>>>>

The goal of this section is to show how to solve RECL constraints by utilizing CEFA. 
At first, we reduce the satisfiability of RECL constraints to a decision problem defined in the sequel. Then we propose a decision procedure for this problem.

\begin{definition}[Nonemptiness of CEFA w.r.t. an LIA formula
, abbreviated as $\cefadec$]
Suppose that $x_1, \cdots, x_n$ are string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, and the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, moreover, $\varphi$ is an LIA formula whose free variables are from $ \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are assignments $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{i,j}}$
such that $(\theta(x_i); \myvec{v_{i,j}}) \in \Lang(\aut_{i, j})$ and  $\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})]$ is true, for every $i \in [n], j \in [l_i]$.
\end{definition}
\begin{proposition}[\cite{atva2020}]\label{prop-cefadec}
$\cefadec$ is PSPACE-complete. 
\end{proposition}
%\begin{proof}
%  The lower bound follows from the fact that the nonemptiness problem of the intersection of NFAs is PSPACE-complete \cite{pspace_int}. For the upper bound, \cite{atva2020} showed that $\cefadec$ is in PSPACE by reducing it to monotonic counter machines.
%\end{proof}
Note that the decision procedure in \cite{atva2020} was only used to prove the upper bound in Proposition~\ref{prop-cefadec} and not implemented as a matter of fact. Instead, the symbolic model checker nuXmv~\cite{nuxmv} was used to solve $\cefadec$. We do not rely on nuXmv in this work and shall propose a new algorithm for solving $\cefadec$ in Section~\ref{subsec:cefadec}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Solving $\cefadec$} \label{subsec:cefadec}

@@

\medskip
\noindent
\emph{Step 1 (Computing intersection automata).} For each $i \in [n]$, compute a CEFA $\cB_{i} = \aut_{i,1} \cap \cdots \cap \aut_{i, l_i}$, and let $\Lambda'_{x_i} := \{\cB_i\}$. \qed

After Step 1, the nonemptiness of CEFAs in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi$ is reduced to the nonemptiness of CEFAs in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$.

In the following steps, we reduce the non-emptiness of CEFAs in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$ to the satisfiability of an LIA formula. The reduction relies on the following two observations. 
%The first observation follows from the fact that $\Lambda'_{x_i}$ for each $i \in [n]$ contains exactly one CEFA, and the second observation follows from the fact that the non-emptiness of exactly one CEFA w.r.t an LIA formula is entirely unrelated to the alphabet.

@@

\begin{observation}\label{obs-cefa-nfa}
For each CEFA $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$, 
%
$$\cefaout(\aut) = \left\{\sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \left \vert\ \eta \in \parikh(\cU_\aut) \mbox{ and }  \alpha\left[  \sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \big{/}  R \right] \mbox{ is true}\right.\right\}.$$
%
\end{observation}

@@

Because the sizes of the LIA formulas $\psi_{\cU_{\cB_1}}(\fZ_{\costset_{\cB_1}}), \cdots, \psi_{\cU_{\cB_n}}(\fZ_{\costset_{\cB_n}})$ are proportional to the sizes (more precisely, the alphabet size, the number of states and transitions) of NFA $\cU_{\cB_1}, \cdots, \cU_{\cB_n}$, and the satisfiability of existential LIA formulas is NP-complete, it is vital to reduce the sizes of these NFAs to improve the performance.

Since
$\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v} = \sum \limits_{\myvec{v} \in \costset(\cB_i) \setminus \{\myvec{0}\}} \eta(\myvec{v}) \myvec{v}$ for each $i \in [n]$ and $\eta \in \parikh(\cU_{\cB_i})$, it turns out that 
the $\myvec{0}$-labeled transitions in $\cU_{\cB_i}$ do not contribute to the final output $\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v}$. Therefore, we can apply the following size-reduction technique for $\cU_{\cB_i}$'s.
%Regardless of alphabet, we take register updates as new alphabet symbols and $\myvec{0}$ as $\epsilon$ symbol. Then we are capable of using standard minimization to reduce the automata size.

\medskip
\noindent
\emph{Step 2 (Reducing automata sizes).} For each $i \in [n]$, we view the transitions $(q, \myvec{0}, q')$ in $\cU_{\cB_i}$ as $\varepsilon$-transitions $(q, \varepsilon, q')$, and remove the $\varepsilon$-transitions from $\cU_{\cB_i}$. Then we determinize and minimize the resulting NFA.   \qed

@@

\medskip
\noindent
\emph{Step 3 (Computing Parikh images).} For each $i \in [n]$, we compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cC_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then we solve the satisfiability of the following formula,
%
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
%\qed
